Higher-Order unification for free!

### 2024-04-15 paper/main.tex
7.1

In this section we suppose the unification of the object language between two
terms $t_1$ and $t_2$ to fail each time at least one of the between $t_1$ or
$t_2$ is outside $\llambda$. This means for instance that $X \nUo Y \appsep Z$ and 
$X\appsep Y \nUo X\appsep Y$.

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
% \noteD[inline]{I've rewritten it, it is clearer?}
In general, unification between \notllambda terms admits more then one
solution and committing one of them in the substitution does not guarantee
\cref{prop:complete-ml}. For instance, $X \appsep a \Uo a$ 
admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
future unifications.

Given a list of unification problems, $\foUnifPb_1\dots
\foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, it is often the case that the resolution of
$\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
that $\rho\foUnifPb_n$ falls again in \llambda.
%
% make test ONLY=7003 TEX=TEX + deactivate beta tex
\printAlll
  {{{X,\lambda x.a},
    {(X\appsep a),a}}}
  {{{A,\lambda x.a},
    {(A\appsep a),a}}}
  {{{X,A,0}}}
  {{}}

In the example above, we see that $\foUnifPb_1$ instantiates $X$ so that
$\foUnifPb_2$ can be solved in \llambda.
On the other hand, we see that, 
\Ue can't solve the compiled problems \hoUnifPb. In
fact, the resolution of $\hoUnifPb_1$ gives the substitution $\sigma = \{ A
\mapsto \lambda x. a\}$, but the dereferencing of $\hoUnifPb_2$ gives the 
non-unifiable problem $(\lambda x. a) \appsep a \nUe a$.

To address
this unification problem, term compilation must recognize and replace
\notllambda terms with fresh variables.
This replacement produces links that we call \linkbeta.

\linkbeta respects \cref{inv:linklhs} and the term on the \rhs has the
following property:

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape \rhsBeta such that $X$ is a
  unification variable with scope $s_1\dots s_n$\footnote{with $s_1\dots s_n$
  that are distinct names} and $t_1\dots t_m$ is a list of terms. This is
  equivalent to \\\elpiIn{app[uva X S | L]}, where \elpiIn{S}~$= s_1\dots s_n$
  and \elpiIn{L}~$=t_1\dots t_m$.
  \label{inv:beta-rhs}
\end{invariant}

@@

Detection of \notllambda is quite simple to implement in the compiler, since it
is sufficient to detect applications with flexible head and argument that
are not in \llambda. The following rule for \notllambda compilation is inserted 
just before rule~\ref{rule:compapp}.

@@

The list \elpiIn{Ag} is split into the list \elpiIn{Pf} and \elpiIn{Extra} such
that \elpiIn{append Pf Extra Ag} and \elpiIn{Pf} is the largest prefix of
\elpiIn{Ag} such that \elpiIn{Pf} is in \llambda. The \rhs of the \linkbeta is
the application of a fresh variable \elpiIn{C} having in scope all the free
variables appearing in the compiled version of \elpiIn{Pf} and \elpiIn{Extra}. The
variable \elpiIn{B}, returned has the compiled term, is a fresh variable having in
scope all the free variables occurring in \elpiIn{Pf1} and \elpiIn{Extra1}.
Note that this construction enforce \cref{inv:beta-rhs}.

@@

\noteD{Si puo togliere se serve spazio}
\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
\end{corollary}

\begin{comment}
\begin{proof}[Proof sketch]
  Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
  \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
  list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
  however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
  no \linkbeta is produced which contradicts our initial assumption.
\end{proof}
\end{comment}

\noteD{Si puo togliere se serve spazio}
\begin{corollary}
  Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
  not a name.
\end{corollary}

\begin{comment}
\begin{proof}[Proof sketch]
  By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
  the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
  $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
  in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
  in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
  \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
\end{proof}
\end{comment}

>>>>

Even if we do not requrie \Ue to handle problems outside of \llambda it is
worth handling terms in \notllambda since, in practice, these terms often
re-enter \llambda at runtime, as observed in~\cite{Nadathur2001} for example,
or to accommodate for the introduction of heuristics.

In the following example problem $\foUnifPb_2$ is $X \appsep a \Uo a$  that
admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
and $\rho_2 = \{X \mapsto \lambda x.a\}$. The unification algorithm alone
cannot chose, since no solution is more general than the other,
although the first problem solve this dilemma at run-time making the second
problem well behaved:
%
\printAlll
  {{{X,\lambda x.a},
    {(X\appsep a),a}}}
  {{{A,\lambda x.a},
    {(A\appsep a),a}}}
  {{{X,A,0}}}
  {{}}

% In this section we suppose the unification of the object language between two
% terms $t_1$ and $t_2$ to fail each time at least one of the between $t_1$ or
% $t_2$ is outside $\llambda$. This means for instance that $X \nUo Y \appsep Z$ and 
% $X\appsep Y \nUo X\appsep Y$.

@@

% \noteE[inline]{IIRC, second order unification, where variables can stand for functions, is semi-decidable, indeed Huet's algorith enumerates all solutions. Third order, functions that take functions, is undecidable, proved by Dowek. IMO here the problem is that there is no unique solution, (semi)decidability is a secondary point. This paragraph needs fixing.}
% \noteD[inline]{I've rewritten it, it is clearer?}
% In general, unification between \notllambda terms admits more then one
% solution and committing one of them in the substitution does not guarantee
% \cref{prop:complete-ml}. For instance, $X \appsep a \Uo a$ 
% admits two different substitutions: $\rho_1 = \{X \mapsto \lambda x.x\}$
% and $\rho_2 = \{X \mapsto \lambda \_.a\}$. Prefer one over the other may break
% future unifications.

% Given a list of unification problems, $\foUnifPb_1\dots
% \foUnifPb_n$ with $\foUnifPb_n$ in \notllambda, it is often the case that the resolution of
% $\bigwedge_{i=0}^{n-1}\foUnifPb_i$ gives a partial substitution $\rho$, such
% that $\rho\foUnifPb_n$ falls again in \llambda.
% %
% % make test ONLY=7003 TEX=TEX + deactivate beta tex
% \printAlll
%   {{{X,\lambda x.a},
%     {(X\appsep a),a}}}
%   {{{A,\lambda x.a},
%     {(A\appsep a),a}}}
%   {{{X,A,0}}}
%   {{}}

\noindent
We want to support this scenario but we have to improve a little our compiler
and \progress routines.
In particular $\hoUnifPb_1$ generates $\sigma = \{ A
\mapsto \lambda x. a\}$ making $\sigma\hoUnifPb_2$ equal to
$(\lambda x. a) \appsep a \Ue a$ that \Ue{} cannot solve since it
lacks rules \ref{clause:beta1} and \ref{clause:beta2}.

To address this problem the compiler must recognize and replace
\notllambda terms with fresh variables and generate a new kind of links
that we call \linkbeta.

In addition to \fullref{inv:linklhs}, the term on the \rhs of a \linkbeta
has the following property:

@@

\begin{invariant}[\linkbeta \rhs]
  The \rhs of any \linkbeta has the shape $X\appsep t_1\dots t_m$
  where $X$ is a unification variable in \llambda (with scope $s_1\dots s_n$)
  and $t_1\dots t_m$ is a list of terms such that $m>0$ and
  $t_1$ is either a variable occurring in $s_1\dots s_n$ or
  a term other than a variable.
  \label{inv:beta-rhs}
\end{invariant}

\noindent Note that the shape of such as \rhs is
  <<\elpiIn{app [uva X S | L]}>>, where \elpiIn{S}~$= s_1\dots s_n$
  and \elpiIn{L}~$=t_1\dots t_m$.

@@

For the detection of \notllambda it
is sufficient to identify applications with a flexible head and arguments that
are not in distinct names. The following rule for \notllambda compilation
is inserted  just before rule~\ref{rule:compapp}.

@@

\noindent
The list \elpiIn{Ag} is split into the two parts: 
\elpiIn{Pf} is in \llambda, and can be empty; 
\elpiIn{Extra} cannot be empty and is such that
\elpiIn{append Pf Extra Ag}.
The \rhs of the \linkbeta is
the application of a fresh variable \elpiIn{C} having
in scope all names in \elpiIn{Pf1} (the compilation of \elpiIn{Pf}).
The variable \elpiIn{B}, returned has the compiled term, is a fresh variable
having in scope all the free variables occurring in \elpiIn{Pf1} and
\elpiIn{Extra1} (the compilation of \elpiIn{Extra}).
This construction enforces \cref{inv:beta-rhs}.

@@

% \noteD{Si puo togliere se serve spazio}
% \begin{corollary}
%   Let \rhsBeta be the \rhs of a \linkbeta, then $m > 0$.
% \end{corollary}

% \begin{comment}
% \begin{proof}[Proof sketch]
%   Assume we have a \linkbeta, by contradiction, if $m = 0$, then the original
%   \Fo{} term has the shape \elpiIn{fapp[fuva M | Ag]} where \elpiIn{Ag} is a
%   list of distinct names (i.e. the list \elpiIn{Extra} is empty). This case is
%   however captured by rule~\ref{rule:complam} (from \cref{sec:compilation}) and
%   no \linkbeta is produced which contradicts our initial assumption.
% \end{proof}
% \end{comment}

% \noteD{Si puo togliere se serve spazio}
% \begin{corollary}
%   Let \rhsBeta be the \rhs of a \linkbeta, then $t_1$ either appears in $s_1\dots s_n$ or it is 
%   not a name.
% \end{corollary}

% \begin{comment}
% \begin{proof}[Proof sketch]
%   By construction, the lists $s_1\dots s_n$ and $t_1\dots t_m$ are built by splitting
%   the list \elpiIn{Ag} from the original term \elpiIn{fapp [fuva A|Ag]}.
%   $s_1\dots s_n$ is the longest prefix of the compiled terms in \elpiIn{Ag} which is
%   in \llambda. Therefore, by definition of \llambda, $t_1$ must appear  
%   in $s_1\dots s_n$, otherwise $s_1\dots s_n$ is not the longest prefix in
%   \llambda, or it is a term with a constructor of \elpiIn{tm} as functor.  
% \end{proof}
% \end{comment}
